---
title: "Research Project 1 (aka mini-project)"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reshape2)
library(stringr)
library(lubridate)
library(ggplot2)

```

# Data Sources and cleaning 

## Sources

These were the data sources used:

  * COVID Data: [UK Coronavirus Dashboard](https://coronavirus.data.gov.uk/ "UK COVID Dashboard")
  * Population: [ONS Population Estimates](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/populationestimatesforukenglandandwalesscotlandandnorthernireland "UK Population Estimates")
  * Lockdown timeline: 
    + [Institute for Government](https://www.instituteforgovernment.org.uk/sites/default/files/timeline-coronavirus-lockdown-december-2021.pdf)
    + [House of Commons library](https://researchbriefings.files.parliament.uk/documents/CBP-9068/CBP-9068.pdf)
  * Temperature: Met Office [Hadley Centre Central England Temperature Data](https://www.metoffice.gov.uk/hadobs/hadcet/data/download.html "HadCET")
  * Rainfall: Met Office Hadley Centre observations datasets: [HadUKP](https://www.metoffice.gov.uk/hadobs/hadukp/ "HadUKP")
  
## Data Cleaning 

### Temperature Data

Data was saved as a `.txt` file. It is stored as follows:

* The first column corresponding to the year
* The second column corresponding to the day
* Each column thereafter corresponding to the month (12 columns, one for each month).

Each cell represents daily values, expressed in tenths of a °C. 

To import data, I created the following function

```{r create.temp.data, echo=TRUE}

clean.temp.data <- function(data, temp.name){
  
  require(reshape2)
  require(lubridate)
  
  temp <- read.fwf(data,
                   widths = rep(5, 14),
                   header = FALSE,
                   col.names = c("Year", "Day",
                                 "Jan", "Feb", "Mar", "Apr",
                                 "May", "Jun", "Jul", "Aug",
                                 "Sep", "Oct", "Nov", "Dec"))
  
  
  # change -999 to NA
  temp[temp == -999] <- NA
  
  # wide to long
  temp <- melt(temp, id = c("Year", "Day"))
  names(temp)[c(3,4)] <- c("Month", temp.name)
  
  # change month to factor
  
  temp$Month <- factor(temp$Month,
                       levels = c("Jan", "Feb", "Mar", "Apr",
                                  "May", "Jun", "Jul", "Aug",
                                  "Sep", "Oct", "Nov", "Dec"),
                       ordered = TRUE)
  
  # introduce decimal points 
  
  temp[[temp.name]] <- (temp[[temp.name]])/10
  
  
  # order data frame by Year, Month and then Day 
  
  temp <- temp[order(temp$Year,
                     temp$Month,
                     temp$Day),]
  
  # Create date variable combining Year, Month and Day 
  
  temp$date <- paste(temp$Year, as.numeric(temp$Month), temp$Day,
                     sep = "-")
  
  temp$date <- ymd(temp$date)
  
  return(temp)
  
}

```

I used this function to import minimum, maximum and mean daily temperatures into a data frame in R.

```{r import_temp_data, echo=TRUE, warning=FALSE}

mean.temp <- clean.temp.data(data = "raw_data/daily_HadCET_1772_2022.txt",
                             temp.name = "mean.temp")

min.temp <- clean.temp.data(data = "raw_data/daily_min_HadCET_1878_2022.txt",
                            temp.name = "min.temp")

max.temp <- clean.temp.data(data = "raw_data/daily_max_HadCET_1878_2022.txt",
                            temp.name = "max.temp")

```

### Rainfall Data

Rainfall is stored in mm/day. The format of the raw data is similar to that of the HadCET data, where:
* First column is Year
* Second column is the Month, stored in its numerical version 
* The rest of the columns correspond to the day, from 1 to 31 

The following code was used to clean up the rainfall data and create a data frame. 

```{r rainfall_data_import, echo=TRUE, warning=FALSE}

rain <- read.fwf("raw_data/daily_rainfall_NWEP_HadUKP.txt", 
                 widths = c(5, 5, rep(7, 31)),
                 header = FALSE)

# rename columns
names(rain) <- c("Year", "Month", seq(1, 31, by = 1))

# change -99.99 to NA

rain[rain==-99.99] <- NA

# wide to long 

rain <- melt(rain, id = c("Year", "Month"))

names(rain)[c(3,4)] <- c("Day", "rain")

# order data by Year, Month and Day 

rain <- rain[order(rain$Year,
                   rain$Month,
                   rain$Day),]

# create date column

rain$date <- paste(rain$Year, rain$Month, rain$Day,
                   sep = "-")
rain$date <- ymd(rain$date)


```

### COVID data

COVID data came fairly well cleaned and needed minimum wrangling. I imported the full data set, containing cases, deaths and vaccinations. I also had a separate dataset with reinfections. (This had to done separately due to the restriction imposed by UK Gov, where a maximum of 5 variables can be downloaded at a time). 

```{r COVID_data_import, echo=TRUE}

covid <- read.csv("raw_data/region_2022-02-24.csv")
reinf <- read.csv("raw_data/reinfections.csv")

covid$date <- ymd(covid$date)
reinf$date <- ymd(reinf$date)

covid <- merge(covid, reinf[,c(4,5)], by = "date")

```

In the vaccination variables, the data for the dates where vaccines had not been rolled out yet had been stored as `NA`. Therefore, to make sure this is not considered as missing data, these have to be changed to 0. 

```{r vaccination_NA, echo=TRUE}

covid$newPeopleVaccinatedSecondDoseByVaccinationDate[is.na(covid$newPeopleVaccinatedSecondDoseByVaccinationDate)] <- 0 

covid$newPeopleVaccinatedThirdInjectionByVaccinationDate[is.na(covid$newPeopleVaccinatedThirdInjectionByVaccinationDate)] <- 0


```

Once this data set was ready, I merged the environmental variables into this data frame. 

```{r merge_env_vars, echo=TRUE}

covid <- merge(covid, rain[,c(4,5)], by = "date")
covid <- merge(covid, mean.temp[,c(4,5)], by = "date")
covid <- merge(covid, min.temp[,c(4,5)], by = "date")
covid <- merge(covid, max.temp[,c(4,5)], by = "date")

```

### Population data

As the ONS only produces a single population estimate per year, and the most recent one corresponds to mid-2020, this was introduced into the `covid` data frame as a constant number across all entries. 

The population for mid-2020 was 7 367 456. 

# Data analysis 

## Exploratory data analysis

### COVID data frame

```{r read_covid_data, echo=FALSE}

covid <- read.csv("data/covid_data.csv")

covid$date <- ymd(covid$date)

```

Before beginning any exploratory data analysis, I had to create a `t` variable. This corresponds to the increase in time. In this case, `t` is day. 

```{r create_t_var, echo=TRUE}

covid$t <- as.numeric(covid$date)-(min(as.numeric(covid$date)))+1

```

I also had to manually create a lockdown variable. This is an indicator variable, where 1 shows current lockdown and 0 shows no lockdown. 

I assumed lockdown was either a full national lockdown or a regional Tier 4 lockdown for the North West. I also assumed that the end of lockdown periods corresponded to either the end of national lockdown or moving into step 1, where stay-at-home orders were raised. 

```{r lockdown_var, echo=TRUE}

covid$lockdown <- rep(0, length(covid$date))

# second lockdown
covid$lockdown[covid$date > ymd("2020-11-04") & 
                 covid$date < ymd("2020-12-02")] <- 1

# third lockdown 
covid$lockdown[covid$date > ymd("2020-12-25") & 
                 covid$date < ymd("2021-03-30")] <- 1

```

### COVID plots

Once these variables were created, I was able to plot the cases of COVID against time. 

```{r COVID_cases_plot, fig.cap="Fig 1 - COVID cases in North West England. Red shading corresponds to lockdown periods.", echo=FALSE}

ggplot(covid, aes(x = t))+
  geom_line(aes(y = cases))+
  annotate("rect", xmin = 132, xmax = 158,
           ymin = -500, ymax = max(covid$cases)+500,
           fill = "red", alpha = 0.1)+
  annotate("rect", xmin = 183, xmax = 276,
           ymin = -500, ymax = max(covid$cases)+500,
           fill = "red", alpha = 0.1)+
  scale_y_continuous("New Cases", expand = c(0,0))+
  xlab("Time (days)")+
  theme_light()

```

It is easier to see trends on the log scale, so I also created a plot using this scale:

```{r log_covid_cases_plot, fig.cap="Fig 2 - COVID Cases for the North West of England, using the log scale. Red shading corresponds to lockdown periods.", echo=FALSE}

ggplot(covid, aes(x = t))+
  geom_line(aes(y = log(cases)))+
  annotate("rect", xmin = 132, xmax = 158,
           ymin = -0.2, ymax = max(log(covid$cases))+0.2,
           fill = "red", alpha = 0.1)+
  annotate("rect", xmin = 183, xmax = 276,
           ymin = -0.2, ymax = max(log(covid$cases))+0.2,
           fill = "red", alpha = 0.1)+
  scale_y_continuous("log(New Cases)", expand = c(0,0))+
  xlab("Time (days)")+
  theme_light()

```

### Create lagged variables 

The next step was to create lagged environmental variables. I built the following function to help me with this:

```{r create.lagged.vars_function, echo=TRUE}

create.lag.var <- function(x,lag,data) {
  
  tlag <- data$t-lag
  
  ind.lag <- sapply(tlag,function(k) {
    
    out <- which((data$t-k)==0)
    
    if(length(out)==0) out <- NA
    
    return(out)
  })
  
  x[ind.lag]
}

```

This function doesn't assume that there is data for every time point. 

I used this function within a `for()` loop to create lags of 2, 7, 10 and 28 days for each of the environmental variables (rainfall, mean temperature, minimum temperature and maximum temperature). 

```{r lag_vars_for_loop, echo=TRUE}

env.vars <- c("rain", "mean.temp", "min.temp", "max.temp")

for (i in 1:length(env.vars)) {
  
  lags <- c(2, 7, 10, 28)
  
  for (j in 1:length(lags)) {
    
    covid[[paste0(env.vars[i], ".lag.", lags[j])]] <- create.lag.var(x = covid[[env.vars[i]]],
                                                                     lag = lags[j],
                                                                     data = covid)
    
  }
  
}


```